The function-sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-08-27 crelle-dictionary.tex
Abstract. Snake.
Returned the abstract to the form we discussed. Also made changes to
the wording of the proof of the main result.

\begin{abstract}
We introduce the category of quasicharacter sheaves on smooth commutative group schemes $G$ over finite fields $k$ and, in doing so, expand the scope of the function-sheaf dictionary from connected commutative algebraic groups to this setting.
We show that the group of isomorphism classes of quasicharacter sheaves is an extension of the group of characters of $G(k)$ by a cohomology group determined by the component group scheme of $G$.
As an application, we find a category of sheaves for which isomorphism classes determine quasi\-characters in the case of algebraic tori, abelian varieties and unipotent wound groups over local fields. 
In this geometrization of quasicharacters, our sheaves are local systems on
Greenberg transforms of locally finite type N\'eron models.
\end{abstract}

@@

\begin{proposition} \label{prop:etale-iso}
 If $G$ is \'etale then the map $S_G : \QCiso{G}\to \Hh^2(E^\bullet_G)$ induces an isomorphism of short exact sequences
\[
\begin{tikzcd}
 0 \arrow{r} & \ker \TrFrob{G} \arrow{d} \arrow{r} & \QCiso{G}\arrow{d}{S_G} \arrow{r}{\TrFrob{G}} \arrow{r} & G(\Fq)^* \arrow{d} \arrow{r} & 0\\
  0 \arrow{r} & \Hh^0(\Weil{},\Hh^2({\bar G},\EEx)) \arrow{r} & \Hh^2(E^\bullet_G) \arrow{r} & \Hh^1(\Weil{},\Hh^1({\bar G},\EEx)) \arrow{r} & 0.
 \end{tikzcd}
 \]
\end{proposition}
\begin{proof}
This result follows easily from Propositions~\ref{prop:ses_desc}, \ref{prop:SGiso} and \ref{prop:sur_etale}.
\end{proof}

Combining Propositions~\ref{prop:sur_etale} and \ref{prop:etale-iso} gives the following

\begin{corollary}\label{cor:split}
The short exact sequence
 \[
    0 \to
    \Hh^0(\Weil{},\Hh^2({\bar G},\EEx)) \to
    \Hh^2(E^\bullet_G) \to
    \Hh^1(\Weil{},\Hh^1({\bar G},\EEx)) \to
    0
 \]
 is split.
\end{corollary}

@@

\begin{proof}
%  Recall the short exact sequence \eqref{eq:pi0}
%  defining the component group scheme for $G$:
%  \[
%  \begin{tikzcd}
%    1 \rar & G^0 \rar & G \rar & \pi_0(G) \rar & 0.
%  \end{tikzcd}
%  \]
  Let
  \[
  \begin{tikzcd}[row sep=20, column sep=20]
%  & 0 \arrow{d} & 0 \arrow{d} & 0\arrow{d} \\
    0 \arrow[dashed]{r} & \ker \TrFrob{\pi_0(G)} \arrow[dashed]{r} \dar & \arrow{d} \ker \TrFrob{G} \arrow[dashed]{r} & \ker \TrFrob{G^0} =0 \arrow[dotted, out=-10, in=170]{dddll} \dar & \\
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\Fq)^* \rar \dar
    & \arrow{d} G(\Fq)^* \rar & G^0(\Fq)^* \rar \dar & 0\\
   &  \coker \TrFrob{\pi_0(G)} =0 \arrow[dashed]{r} & \coker \TrFrob{G} \arrow[dashed]{r} & \coker \TrFrob{G^0} =0 & 
%    & \arrow{d} \coker \TrFrob{\pi_0(G)} =0 \arrow[dashed]{r} & \arrow{d} \coker \TrFrob{G} \arrow[dashed]{r} &  \arrow{d} \coker \TrFrob{G^0} =0 & \\
%      & 0 & 0 & 0 \\
  \end{tikzcd}
  \]
  be the commutative diagram of abelian groups obtained by applying
  Lemma~\ref{lem:pullback} to \eqref{eq:pi0}.
 %
  The sequence of abelian groups
  \begin{equation}\label{eq:pi0k}
  \begin{tikzcd}
    1 \rar & G^0(\Fq) \rar & G(\Fq) \rar & \pi_0(G)(\Fq) \rar & 0,
  \end{tikzcd}
  \end{equation}
  is exact since $\Hh^1(\Fq,G^0) =0$ by Lemma~\ref{lem:G0alg-grp} and Lang's theorem on connected algebraic groups over finite fields \cite{lang:56a}.
  Since $\EEx$ is divisible, $\Hom(\ - \ ,\EEx)$ is exact and thus the dual sequence of
  character groups is also exact.
%
  The upper row is exact by Proposition~\ref{prop:middleexact}.
  Now Lemma~\ref{lem:G0alg-grp} and Proposition~\ref{prop:connected}
  imply that $\ker \TrFrob{G^0} =0$ and $\coker \TrFrob{G^0}=0$,
  while Proposition~\ref{prop:etale-iso} gives $\ker \TrFrob{\pi_0(G)} \iso \Hh^0(\Weil{},\Hh^2(\pi_0({\bar G}),\EEx))$
  and $\coker \TrFrob{\pi_0(G)}=0$.
%
It follows from the snake lemma that $\coker \TrFrob{G} =0$
and $\ker \TrFrob{\pi_0(G)} \to \ker \TrFrob{G}$ is an isomorphism.
This calculates $\ker  \TrFrob{G}$ and shows that
\[
  \begin{tikzcd}
0 \arrow{r} & \Hh^0(\Weil{},\Hh^2(\pi_0(\bG),\EEx)) \arrow{r} & \QCiso{G} \arrow{r}{\TrFrob{G}} & G(\Fq)^* \arrow{r} & 0 
  \end{tikzcd}
\]
is exact.
\end{proof}

>>>>

\begin{abstract}
We introduce the category of quasicharacter sheaves on smooth commutative group schemes $G$ over finite fields $k$ and expand the scope of the function-sheaf dictionary from connected commutative algebraic groups to this setting.
We show that the group of isomorphism classes of quasicharacter sheaves is an extension of the group of characters of $G(k)$ by a cohomology group determined by the component group scheme of $G$.
As an application, we find a category of sheaves for which isomorphism classes correspond to quasi\-characters in the case of algebraic tori, abelian varieties and unipotent wound groups over local fields. 
In this geometrization of quasicharacters, our sheaves are local systems on
Greenberg transforms of locally finite type N\'eron models.
\end{abstract}

@@

\begin{proposition} \label{prop:etale-iso}
 If $G$ is \'etale then the map $S_G : \QCiso{G}\to \Hh^2(E^\bullet_G)$ induces an isomorphism of split short exact sequences
\[
\begin{tikzcd}
 0 \arrow{r} & \ker \TrFrob{G} \arrow{d} \arrow{r} & \QCiso{G}\arrow{d}{S_G} \arrow{r}{\TrFrob{G}} \arrow{r} & G(\Fq)^* \arrow{d} \arrow{r} & 0\\
  0 \arrow{r} & \Hh^0(\Weil{},\Hh^2({\bar G},\EEx)) \arrow{r} & \Hh^2(E^\bullet_G) \arrow{r} & \Hh^1(\Weil{},\Hh^1({\bar G},\EEx)) \arrow{r} & 0.
 \end{tikzcd}
 \]
\end{proposition}
\begin{proof}
This result follows easily from Propositions~\ref{prop:ses_desc}, \ref{prop:SGiso} and \ref{prop:sur_etale}.
\end{proof}

@@

\begin{proof}
%  Recall the short exact sequence \eqref{eq:pi0}
%  defining the component group scheme for $G$:
%  \[
%  \begin{tikzcd}
%    1 \rar & G^0 \rar & G \rar & \pi_0(G) \rar & 0.
%  \end{tikzcd}
%  \]
  Let
  \begin{equation}\label{eq:pre-snake}
  \begin{tikzcd}[row sep=20, column sep=20]
%  & 0 \arrow{d} & 0 \arrow{d} & 0\arrow{d} \\
%    0 \arrow[dotted]{r} & \ker \TrFrob{\pi_0(G)} \arrow[dotted]{r} \dar & \arrow{d} \ker \TrFrob{G} \arrow[dotted]{r} & \ker \TrFrob{G^0} =0 \arrow[dotted, out=-10, in=170]{dddll} \dar & \\
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\Fq)^* \rar
    &  G(\Fq)^* \rar & G^0(\Fq)^* \rar & 0
%   &  \coker \TrFrob{\pi_0(G)} =0 \arrow[dotted]{r} & \coker \TrFrob{G} \arrow[dotted]{r} & \coker \TrFrob{G^0} =0 & 
%    & \arrow{d} \coker \TrFrob{\pi_0(G)} =0 \arrow[dashed]{r} & \arrow{d} \coker \TrFrob{G} \arrow[dashed]{r} &  \arrow{d} \coker \TrFrob{G^0} =0 & \\
%      & 0 & 0 & 0 \\
  \end{tikzcd}
  \end{equation}
  be the commutative diagram of abelian groups obtained by applying
  Lemma~\ref{lem:pullback} to \eqref{eq:pi0}.
 %
  The sequence of abelian groups
\[
  \begin{tikzcd}
    1 \rar & G^0(\Fq) \rar & G(\Fq) \rar & \pi_0(G)(\Fq) \rar & 0,
  \end{tikzcd}
\]
  is exact since $\Hh^1(\Fq,G^0) =0$ by Lemma~\ref{lem:G0alg-grp} and Lang's theorem on connected algebraic groups over finite fields \cite{lang:56a}.
  Since $\EEx$ is divisible, $\Hom(\ - \ ,\EEx)$ is exact and thus the dual sequence of
  character groups in \eqref{eq:pre-snake} is exact.
%
  The upper row in \eqref{eq:pre-snake} is exact by Proposition~\ref{prop:middleexact}.
  Now Lemma~\ref{lem:G0alg-grp} and Proposition~\ref{prop:connected}
  imply that $\ker \TrFrob{G^0} =0$ and $\coker \TrFrob{G^0}=0$,
  while Proposition~\ref{prop:etale-iso} gives $\ker \TrFrob{\pi_0(G)} \iso \Hh^0(\Weil{},\Hh^2(\pi_0({\bar G}),\EEx))$
  and $\coker \TrFrob{\pi_0(G)}=0$.
It now follows from the snake lemma 
   %\begin{equation}\label{eq:snake}
 \[
  \begin{tikzcd}[row sep=20, column sep=15]
%  & 0 \arrow{d} & 0 \arrow{d} & 0\arrow{d} \\
    0 \arrow[dotted]{r} & \ker \TrFrob{\pi_0(G)} \arrow[dotted]{r} \dar & \arrow{d} \ker \TrFrob{G} \arrow[dotted]{r} & \ker \TrFrob{G^0} =0 \arrow[dotted, out=-10, in=170]{dddll} \dar & \\
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\Fq)^* \rar \dar
    & \arrow{d} G(\Fq)^* \rar & G^0(\Fq)^* \rar \dar & 0\\
   &  \coker \TrFrob{\pi_0(G)} =0 \arrow[dotted]{r} & \coker \TrFrob{G} \arrow[dotted]{r} & \coker \TrFrob{G^0} =0 & 
%    & \arrow{d} \coker \TrFrob{\pi_0(G)} =0 \arrow[dashed]{r} & \arrow{d} \coker \TrFrob{G} \arrow[dashed]{r} &  \arrow{d} \coker \TrFrob{G^0} =0 & \\
%      & 0 & 0 & 0 \\
  \end{tikzcd}
  \]
 % \end{equation}
that $\coker \TrFrob{G} =0$
and $\ker \TrFrob{\pi_0(G)} \to \ker \TrFrob{G}$ is an isomorphism.
This gives the short exact sequence
\[
  \begin{tikzcd}
0 \arrow{r} & \Hh^0(\Weil{},\Hh^2(\pi_0(\bG),\EEx)) \arrow{r} & \QCiso{G} \arrow{r}{\TrFrob{G}} & G(\Fq)^* \arrow{r} & 0. 
  \end{tikzcd}
\]
\end{proof}
